---
title: "Pattern break experiments"
author: "Othman El Hammouchi"
---

```{r}
library(tidyverse)
library(ggplot2)
library(magrittr)
library(ggfortify)
library(ChainLadder)
library(patchwork)
library(MASS)
library(knitr)
```

```{r}
bootReserve <- function(inputTriangle, nBoot) {

    nDev <- ncol(inputTriangle)

    mackResults <- suppressWarnings(MackChainLadder(inputTriangle))

    devFacs <- mackResults$f[-nDev]
    sigma <- mackResults$sigma

    F_list <- list()
    prevCs <- list()
    resids <- list()

    for (idx in 1:(nDev - 1)) {
        F_list[[idx]] <- unname(inputTriangle[1:(nDev - idx), idx + 1] / inputTriangle[1:(nDev - idx), idx])
        prevCs[[idx]] <- unname(inputTriangle[1:(nDev - idx), idx])
        resids[[idx]] <- (F_list[[idx]] - devFacs[idx]) * sqrt(prevCs[[idx]]) / sigma[idx]
    }

    #sample residuals and put them in proper list format
    reserveBoot <- c()
    for (iBoot in 1:nBoot) {
        residBoot <- list()
        residSample <- sample(unlist(resids), replace = TRUE)
        for (j in 1:(nDev - 1)) {
            residBoot[[j]] <- residSample[1:(nDev - j)]
            residSample <- residSample[-1:-(nDev - j)]
        }

        #compute bootstrapped quantities
        FBoot <- list()
        fBoot <- c()
        sigmaBoot <- c()
        for (j in 1:(nDev - 1)) {
            FBoot[[j]] <- devFacs[j] + (residBoot[[j]] * sigma[j]) / sqrt(prevCs[[j]])
            fBoot[j] <- sum(FBoot[[j]] * prevCs[[j]]) / sum(prevCs[[j]])
            sigmaBoot[j] <- mean(prevCs[[j]] * (FBoot[[j]] - fBoot[j])**2)
        }

        #complete the triangle
        bootTriangle <- inputTriangle
        for (diagIdx in 1:(nDev-1)) {
            for (rowIdx in (diagIdx + 1):nDev){
                colIdx <- nDev + diagIdx + 1 - rowIdx
                bootTriangle[rowIdx, colIdx] <-
                    rnorm(
                        1,
                        bootTriangle[rowIdx, colIdx - 1] * fBoot[colIdx - 1],
                        sqrt(bootTriangle[rowIdx, colIdx - 1]) * sigmaBoot[colIdx - 1]
                    )
                }
            }
        #if the triangle contains undefined values, the simulated reserve became negative at some point, and we throw away the triangle
        if (!(NA %in% bootTriangle) && !(NaN %in% bootTriangle)) {
            latest <- bootTriangle[col(bootTriangle) + row(bootTriangle) == nDev + 1]
            reserve <- sum(bootTriangle[, nDev] - latest)
            reserveBoot <- c(reserveBoot, reserve)
        }
    }
    return(reserveBoot)
}
```

We start by generating a claims triangle which satisfies the assumptions of Mack's method.

```{r}
nDev <- ncol(UKMotor)
benchmark <- MackChainLadder(UKMotor)
sigma <- benchmark$sigma
devFac <- benchmark$f[-nDev]
initMean <- mean(UKMotor[, 1])
initStd <- sd(UKMotor[, 1])
```

```{r}
initCol <- rnorm(nDev, initMean, initStd)

claimsTriangle <- matrix(ncol = nDev, nrow = nDev)
claimsTriangle[, 1] <- initCol

for (colIdx in 2:nDev) {
    for (rowIdx in 1:(nDev + 1 - colIdx)) {
        prevC <- claimsTriangle[rowIdx, colIdx - 1]
        claimsTriangle[rowIdx, colIdx] <-
            rnorm(1, devFac[colIdx - 1] * prevC, sigma[colIdx - 1] * sqrt(prevC))
    }
}

# kable(claimsTriangle)
```

As expected, the Mack estimator performs very well in this case:

```{r}
res <- MackChainLadder(claimsTriangle)

print(res$f)
```

\section{Perturbing a single observation}

First, we study the effect of perturbing a single observation and propagating this forward. The following function generates such data given a desired outlier column and row and a perturbation factor. 

```{r}
singleOutlier <- function(nDev, initMean, initStd, devFac, sigma, outlierColIdx, outlierRowIdx, pert = 1.1) {
    initCol <- rnorm(nDev, initMean, initStd)

    claimsTriangle <- matrix(ncol = nDev, nrow = nDev)
    claimsTriangle[, 1] <- initCol

    for (colIdx in 2:nDev) {
        for (rowIdx in 1:(nDev + 1 - colIdx)) {
            prevC <- claimsTriangle[rowIdx, colIdx - 1]
            claimsTriangle[rowIdx, colIdx] <-
                rnorm(1, devFac[colIdx - 1] * prevC, sigma[colIdx - 1] * sqrt(prevC))
        }
    }
    claimsTriangle[outlierRowIdx, -1] <- NA

    if (outlierColIdx > 2) {
        for (colIdx in 2:(outlierColIdx - 1)) {
            prevC <- claimsTriangle[outlierRowIdx, colIdx - 1]
            claimsTriangle[outlierRowIdx, colIdx] <-
                rnorm(1, devFac[colIdx - 1] * prevC, sigma[colIdx - 1] * sqrt(prevC))
        }
    }
    prevC <- claimsTriangle[outlierRowIdx, outlierColIdx - 1]
    claimsTriangle[outlierRowIdx, outlierColIdx] <-
        rnorm(1,
            pert * devFac[outlierColIdx - 1] * prevC,
            sigma[outlierColIdx - 1] * sqrt(prevC))

    if (outlierColIdx < nDev) {
        for (colIdx in (outlierColIdx + 1):(nDev + 1 - outlierRowIdx)) {
            prevC <- claimsTriangle[outlierRowIdx, colIdx - 1]
            claimsTriangle[outlierRowIdx, colIdx] <-
                rnorm(1, devFac[colIdx - 1] * prevC, sigma[colIdx - 1] * sqrt(prevC))
        }
    }
    return(claimsTriangle)
}
```

We can repeat the simulation for every possible point and a range of perturbations between 0.5 and 1.5. 

```{r}
results <- tibble(
    row = numeric(),
    col = numeric(),
    pert = numeric(),
    triangle = list(),
    reserve = list()
)

for (colIdx in 2:nDev) {
    for (rowIdx in 1:(nDev + 1 - colIdx)) {
        for (pert in seq(0.5, 1.5, length.out = 10)) {
            triangle <- singleOutlier(nDev, initMean, initStd, devFac, sigma, colIdx, rowIdx, pert = pert)
            reserve <- bootReserve(triangle, 1e3) %>% list()
            triangle %<>% list()
            results %<>% add_row(row = rowIdx, col = colIdx, triangle = triangle, pert = pert, reserve = reserve)
        }
    }
}
```

Now let's take a look at the bootstrap mean as a function of the perturbation for different rows and columns.

```{r}
results %<>%
    mutate(mean = map(reserve, mean, na.rm = TRUE)) %>%
    unnest_auto(mean)
```

```{r}
ggplot(data = results) +
    geom_point(aes(pert, mean)) +
    facet_wrap(vars(col))
```

```{r}
ggplot(data = results) +
    geom_point(aes(pert, mean)) +
    facet_wrap(vars(row))
```

Surprisingly, the plots show a higher mean both for smaller and larger perturbations.

\subsection{Gamma distribution}

The problem with using the above normal model is that it allows for negative samples during the bootstrap simulation. To avoid this, we can use a different distribution which still respects Mack's assumptions. If we take for instance $C_{ij} \sim \Gamma(\alpha, \beta)$, then we must choose $\alpha, \beta$ to satisfy

$$
\begin{cases}
\frac{\alpha}{\beta} = f_{j-1} C_{i, j-1} \\
\frac{\alpha}{\beta^2} = \sigma^2_{j-1} C_{i, j-1} \,,
\end{cases}
$$

giving the values

$$
\alpha = \frac{f_{j-1}^2 C_{i, j-1}}{\sigma_{j-1}^2} \\
\beta = \frac{f_{j-1}}{\sigma_{j-1}^2} \,,
$$

for the distribution parameters.

```{r}
bootReserveGamma <- function(inputTriangle, nBoot) {

    nDev <- ncol(inputTriangle)

    mackResults <- suppressWarnings(MackChainLadder(inputTriangle))

    devFacs <- mackResults$f[-nDev]
    sigma <- mackResults$sigma

    F_list <- list()
    prevCs <- list()
    resids <- list()

    for (idx in 1:(nDev - 1)) {
        F_list[[idx]] <- unname(inputTriangle[1:(nDev - idx), idx + 1] / inputTriangle[1:(nDev - idx), idx])
        prevCs[[idx]] <- unname(inputTriangle[1:(nDev - idx), idx])

        mean <- devFacs[idx - 1] ** 2 * prevCs[[idx]] / sigma[idx - 1] ** 2
        std <- devFac[idx - 1] * prevCs[[idx]] / sigma[idx - 1] ** 2

        resids[[idx]] <- (F_list[[idx]] - mean) / std
    }

    #sample residuals and put them in proper list format
    reserveBoot <- c()
    for (iBoot in 1:nBoot) {
        residBoot <- list()
        residSample <- sample(unlist(resids), replace = TRUE)
        for (j in 1:(nDev - 1)) {
            residBoot[[j]] <- residSample[1:(nDev - j)]
            residSample <- residSample[-1:-(nDev - j)]
        }

        #compute bootstrapped quantities
        FBoot <- list()
        fBoot <- c()
        sigmaBoot <- c()
        for (j in 1:(nDev - 1)) {
            mean <- devFacs[idx - 1] ** 2 * prevCs[[idx]] / sigma[idx - 1] ** 2
            std <- devFac[idx - 1] * prevCs[[idx]] / sigma[idx - 1]**2

            FBoot[[j]] <- residBoot[[j]] * std + mean
            fBoot[j] <- sum(FBoot[[j]] * prevCs[[j]]) / sum(prevCs[[j]])
            sigmaBoot[j] <- mean(prevCs[[j]] * (FBoot[[j]] - fBoot[j])**2)
        }

        #complete the triangle
        bootTriangle <- inputTriangle
        for (diagIdx in 1:(nDev-1)) {
            for (rowIdx in (diagIdx + 1):nDev){
                colIdx <- nDev + diagIdx + 1 - rowIdx
                bootTriangle[rowIdx, colIdx] <-
                    rgamma(1,
                    devFac[colIdx - 1]**2 * bootTriangle[rowIdx, colIdx - 1] / sigma[colIdx - 1]**2,
                    devFac[colIdx - 1] / (sigma[colIdx - 1]**2))
                }
            }
        #if the triangle contains undefined values, the simulated reserve became negative at some point, and we throw away the triangle
        if (!(NA %in% bootTriangle) && !(NaN %in% bootTriangle)) {
            latest <- bootTriangle[col(bootTriangle) + row(bootTriangle) == nDev + 1]
            reserve <- sum(bootTriangle[, nDev] - latest)
            reserveBoot <- c(reserveBoot, reserve)
        }
    }
    return(reserveBoot)
}
```

```{r}
singleOutlierGamma <- function(nDev, initMean, initStd, devFac, sigma, outlierColIdx, outlierRowIdx, pert = 1.1) {

    initCol <- rgamma(nDev, initMean**2 / initStd**2, initMean / initStd**2)

    claimsTriangle <- matrix(ncol = nDev, nrow = nDev)
    claimsTriangle[, 1] <- initCol

    for (colIdx in 2:nDev) {
        for (rowIdx in 1:(nDev + 1 - colIdx)) {
            prevC <- claimsTriangle[rowIdx, colIdx - 1]
            claimsTriangle[rowIdx, colIdx] <-
                rgamma(1,
                    devFac[colIdx - 1]**2 * claimsTriangle[rowIdx, colIdx - 1] / sigma[colIdx - 1]**2,
                    devFac[colIdx - 1] / (sigma[colIdx - 1]**2))
        }
    }

    claimsTriangle[outlierRowIdx, -1] <- NA

    if (outlierColIdx > 2) {
        for (colIdx in 2:(outlierColIdx - 1)) {
            prevC <- claimsTriangle[outlierRowIdx, colIdx - 1]
            claimsTriangle[outlierRowIdx, colIdx] <-
                rgamma(1,
                    devFac[colIdx - 1]**2 * prevC / sigma[colIdx - 1]**2,
                    devFac[colIdx - 1] / sigma[colIdx - 1]**2)
        }
    }

    prevC <- claimsTriangle[outlierRowIdx, outlierColIdx - 1]
    claimsTriangle[outlierRowIdx, outlierColIdx] <-
        rgamma(1,
            (pert * devFac[outlierColIdx - 1])**2 * prevC / sigma[outlierColIdx - 1]**2,
            (pert * devFac[outlierColIdx - 1]) / sigma[outlierColIdx - 1]**2)

    if (outlierColIdx < nDev) {
        for (colIdx in (outlierColIdx + 1):(nDev + 1 - outlierRowIdx)) {
            prevC <- claimsTriangle[outlierRowIdx, colIdx - 1]
            claimsTriangle[outlierRowIdx, colIdx] <-
                rgamma(1,
                    devFac[colIdx - 1]**2 * prevC / sigma[colIdx - 1]**2,
                    devFac[colIdx - 1] / sigma[colIdx - 1]**2)
        }
    }
    return(claimsTriangle)
}
```

```{r}
results <- tibble(
    row = numeric(),
    col = numeric(),
    pert = numeric(),
    triangle = list(),
    reserve = list()
)

for (colIdx in 2:nDev) {
    for (rowIdx in 1:(nDev + 1 - colIdx)) {
        for (pert in seq(0.5, 1.5, length.out = 10)) {
            triangle <- singleOutlierGamma(nDev, initMean, initStd, devFac, sigma, colIdx, rowIdx, pert = pert)
            reserve <- bootReserveGamma(triangle, 1e3) %>% list()
            triangle %<>% list()
            results %<>% add_row(row = rowIdx, col = colIdx, triangle = triangle, pert = pert, reserve = reserve)
        }
    }
}
```

```{r}
results %<>%
    mutate(mean = map(reserve, mean, na.rm = TRUE)) %>%
    unnest_auto(mean)
```

```{r}
ggplot(data = results) +
    geom_point(aes(pert, mean)) +
    facet_wrap(vars(col))
```

```{r}
ggplot(data = results) +
    geom_point(aes(pert, mean)) +
    facet_wrap(vars(row))
```

